Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-14 application.tex
updating

\section{Application to protoplanetary disks}\label{application} 
As an application of our linear models, we estimate where in a
protoplanetary disk do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one can consider the energy equation with
radiative cooling, 
\begin{align}\label{real_energy}
%\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
  c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
 \end{align}
%where the temperature $T$ is defined through $P=\mathcal{R}\rho
% T/\mu$, $\mathcal{R}$ is the ideal gas constant and $\mu$ is
% the mean molecular weight, 
where $c_v$ is the heat capacity at constant
volume, 
% In Eq. \ref{real_energy} $S$ is the
% specific entropy,
% \begin{align}
%   S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
% \end{align}
% where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
% specific heat capcity at constant volume
$\bm{F}$ is the radiative flux and $Q_+$ is a heating
term such that an equilibrium solution exists.

One should now model in detail each of the terms on the RHS of
Eq. \ref{real_energy}, linearize them, then solve 
the resulting linear eigenvalue problem. Such an approach is beyond 
the scope of this paper. Here, we simply model the RHS of 
Eq. \ref{real_energy} such that it can be represented by a 
thermal relaxation time $t_c$ for the disk temperature. 
%Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
%the 
To do so, we first consider separately perturbations
with radial lengthscales $l\gg l_\mathrm{rad}$ and 
$l\ll l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

%need to comment on kappa_d(rho,T)

\subsection{Radiative diffusion}
For perturbations with lengthscales $l\gg
l_\mathrm{rad}$, we assume  
temperature fluctuations are smoothed out by radiative diffusion. We
model this in linear theory by writing the linearized RHS of
Eq. \ref{real_energy} as 
% we choose the heating term such that
% \begin{align}
%   Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
%     \left(T-T_{t=0}\right)\right], 
% \end{align} 
% where $k_\mathrm{rad}$ is the conduction coefficient to be defined
% implicitly. Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
%     c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
% \end{align}
% When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

@@

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin. In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
flucations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
% the radiative cooling is given by {\bf (reference?)}
% \begin{align}
% \nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
% \end{align}   
% where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
% heating term 
% \begin{align}
%   Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
% \end{align}
% Then Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
%   \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right), 
% \end{align}
% where $c_v$ is the heat capacity at constant volume. When linearized, the last term on the RHS becomes
% \begin{align}
%   -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
%     \frac{P}{\rho}\delta\rho\right). 
% \end{align}
% By comparing this expression with the linearized version of the energy
% equation used in our calculations (Eq. \ref{full_energy1}), we identify 
% \begin{align}\label{tc_newton_cool} 
%   t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
% \end{align}
% as the thermal relaxation timescale for optically thin cooling. 
\begin{align}
  t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}
We note that $t_\mathrm{cool}$ does not explicitly depend on $\rho$.

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
% where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
% $\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
% and Eq. \ref{eta_def}. 
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the 
% In order to utilize our linear results, which assumes $t_c$ is a
% constant, we need to make further 
% approximations. Specifically, we consider vertically-isothermal disks
% and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
% H$, where $\Sigma$ is the surface density at the fiducial
% radius of interest. For example, the dimensionless diffusion
% coefficient becomes 
% \begin{align}\label{diff_coeff_dimensionless}
%   \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
%     c_v\Omega_k}. 
% \end{align}
% Furthermore, we will choose a temperature-dependent opacity model for
% simplicity, so that $\kappa_d=\kappa_d(T)$.
dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{\Sigma^2\Omega_k}{\eta\rho^2}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that the thermal relaxation time depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane. %The latter is due to the density-dependence in the diffusion time. 
%is then a constant for vertically-isothermal disks, 
%and we can directly apply previous results.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
  \beta(z; \khat) =
  &\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\notag\\
&\times\left[\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right].        
\end{align} 
We refer to the case with $\gamma=1.4$, $\mu=2.33$ and the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN) and $\beta=\beta_\mathrm{MMSN}$, 
\begin{align}\label{beta_mmsn_simp}
  \beta_\mathrm{MMSN} = 3.9\times10^{-3} r_\mathrm{AU}^{9/14}\left( 1 +
    1.9\times10^7\hat{\rho}^2r_\mathrm{AU}^{-33/7}\khat^{-2}\right). 
\end{align}
% where
% \begin{align}
%   \
% %  \frac{3.7\times10^{-3}}{\mu\left(\gamma-1\right)\hat{\kappa}_d\hat{T}^5}, 
% \end{align}
% and
% \begin{align}
%   \khat_0 = 4.4\times10^{3}\hat{\kappa}_d\hat{T}^2\hat{\Sigma}.
% \end{align}

% The physical thermal relaxation timescale $\beta$ is height-dependent,
% whilst for analytical discussion we assumed it is
% constant. Nevertheless,
It is instructive to compare the above thermal timescales
with the criterion Eq. \ref{iso_vsi_cond} evaluated for the fiducial disk model,    
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is needed for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
% becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ $\gamma=1.4$, $\hat{T}=\hat{\Sigma}=1$ for three
opacity scales: $\hat{\kappa}_d=0.1, \,1 $ and $10$ at two heights:
$z=0$ and $z=3H$. We also plot Eq. \ref{bcrit_mmsn} in each panel.

Consider first the MMSN shown in the middle panels of 
Fig. \ref{mmsn_bcrit_bcool}. The critical thermal timescale,
Eq. \ref{bcrit_mmsn}, as derived from our analytical discussion, is
most useful to assess the VSI in the optically thin regime (where the
curves overlap), since in this limit $\beta$ does not depend on
height. We thus infer that the VSI can operate at tens of AU in the
MMSN.

In the inner few AU of the MMSN, however, $\beta$ depends on $z$. 
Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
the optically thick midplane, but with increasing height the disk 
becomes optically thin and $\beta<\beta_\mathrm{crit}$. Our analytical
criteria is not valid for variable $\beta$, and we must turn to
numerical solutions of the linear problem to determine whether or not
the VSI operates. We do this in the next section.

Now consider lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}. The curves shift inward, but are also
raised. In fact, lowering the opacity by an order of magnitude renders 
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This is
expected to stabilize the disk. Raising the opacity, as shown in 
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the instability is
restricted to larger radii.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
by choosing a sufficiently large $\khat$. We
illustrate this in Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at
each radius the value of $\khat$ such 
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI,
should it develop in the inner disk, would only occur on very small
scales. However, our linear calculations suggest growth rates
eventually decay for large $\khat$.

Furthermore, modes with large $\khat$ may be subject to viscous decay
in a real disk. The viscous time for a perturbation lengthscale
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha_s\khat^2\Omega $, where
$\alpha_s$ is the standard alpha viscosity parameter
\citep{shakura73}. Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/\epsilon\Omega$ places an upper limit
$\khat^2\lesssim \epsilon/\alpha_s$. Inserting $\epsilon \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha_s\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha_s\sim
10^{-5}$. \citep[These low viscosity levels are required for the
VSI, see][]{nelson13}.

% However, our linear
% calculations indicate growth rates eventually decay as one increases
% $\khat$. This, together with our finding of high $\khat$ modes may be
% affected by vertical boundary conditions, suggest the VSI does not
% operate in the inner few AU in protoplanetary disks where thermal
% relaxation is entirely due to radiation \citep[cf.][]{stoll14}.

@@

\subsection{VSI in the MMSN}
Here we solve the linear stability problem in the MMSN with the scale and
height dependent thermal relaxation function defined by
Eq. \ref{beta_mmsn_simp}.  We use a vertical
domain $z_\mathrm{max}=5H$. In Fig. \ref{mmsn_overall} we plot growth
timescales $t_\mathrm{grow} \equiv 1/\nu$ associated with the
fundamental mode in units of the local orbital time
$P_\mathrm{orb}\equiv 2\pi/\Omega_k$.

In fact, for $1\lesssim \khat \lesssim 30$ we either only find the
fundamental mode, or that it is the most unstable. The associated 
growth rates are also insensitive to $z_\mathrm{max}$. However, for
large $\khat$ and/or $z_\mathrm{max}$, e.g. $\khat=50$ or 
$z_\mathrm{max}=10H$, we also found modes with shorter growth times 
than those shown in Fig. \ref{mmsn_overall}, but these disturbances are 
concentrated near the vertical boundaries
\citep{nelson13,mcnally14}. Such modes may not be robust because    
they are affected by boundary conditions and viscosity in a real
disk. % , and high $\khat$ modes are % subject to viscous decay in a real disk.

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits. 
Importantly, it is necessary to consider smaller scales as $r$
decreases as the disk generally becomes optically thick.

Fig. \ref{mmsn_overall} is roughly consistent with the 
mid-plane thermal relaxation timescales shown in
Fig. \ref{mmsn_bcrit_bcool}. For example, for $\khat=10$ we find 
$\beta \gtrsim \beta_\mathrm{crit}$ for $r\lesssim 10$AU at $z=0$; and
Fig. \ref{mmsn_overall} show stabilization inside $\sim 8$AU where
$t_\mathrm{grow}\to\infty$.

%The mode % operates interior to the expected minimum radius, 
% probably because the disk bcomes optically thin away from the
% midplane.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the fundamental VSI mode in
    the MMSN. These are calculated using the scale and height
    dependent thermal relaxation given by Eq. \ref{beta_mmsn_simp}. 
    \label{mmsn_overall}}    
\end{figure}

@@

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with much
reduced growth rates (when scaled by the local 
rotation).  
%: Fig. \ref{mmsn_overall}) show that
%the growth time at

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer at $r=5$AU than at $r=50$AU for which  $\beta <
\beta_\mathrm{crit}  \forall z$: we find $t_\mathrm{grow}\sim 40$
orbits at $r=50$AU, but $t_\mathrm{grow}\sim 400$ orbits at
$r=5$AU. Thus, introducing an optically thick midplane significantly
reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

% The grow
% mode because it involves vertical motion across $z=0$, which is

% the case with $\beta < \beta_\mathrm{crit}$ for all $z$.

% Introducing an optically thick mid-plane may not completely surpress
% the VSI because the vertical entropy gradient, which usually provides
% stabilization, vanishes at $z=0$. (Note that $N_z^2\propto z^2$ while
% vertical shear $d\Omega^2/dz \propto z$)

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_5AU}
  \caption{Vertical velocity perturbations of the VSI in the MMSN at 
    $r=50$AU (top) and $r=5$AU (bottom). These are calculated using the height
    dependent thermal relaxation given by Eq. \ref{beta_mmsn_simp}. The real (imaginary) part of 
    $\delta v_z$ is shown as
    the solid (dotted) curves. % The eigenfunction has been normalized
    % such that $\imag(\delta v_z)=0$ at the vertical boundaries, and
    % $\mathrm{max}|\delta v_z|=1$. 
    \label{mmsn_eigenvz}}
\end{figure}

>>>>

\section{Application to protoplanetary disks}\label{application} 
As an application of our linear models, we estimate where in a
protoplanetary disk do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one can consider the energy equation
with radiative cooling, 
\begin{align}\label{real_energy}
  % \rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
  \frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
    c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where $c_v$ is the heat capacity at constant
volume, 
$\bm{F}$ is the radiative flux and $Q_+$ is a heating
term to permit an equilibrium solution.

One should now model in detail each of the terms on the RHS of
Eq. \ref{real_energy}, linearize them, then solve 
the resulting linear eigenvalue problem. This is beyond 
the scope of the present work. Here, we take the simpler approach of
modeling the linearized RHS of Eq. \ref{real_energy} such that it can 
be represented by a thermal relaxation time $t_c$. 
To do so, we first consider separately perturbations
with radial lengthscales $l\gg l_\mathrm{rad}$ and  
$l\ll l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

\subsection{Radiative diffusion}
For perturbations with lengthscales $l\gg
l_\mathrm{rad}$, we assume temperature fluctuations are smoothed out
by radiative diffusion and model this in linear theory by writing the
linearized RHS of Eq. \ref{real_energy} as 
% we choose the heating term such that
% \begin{align}
%   Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
%     \left(T-T_{t=0}\right)\right], 
% \end{align} 
% where $k_\mathrm{rad}$ is the conduction coefficient to be defined
% implicitly. Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
%   c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
% \end{align}
% When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

@@

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin. In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
flucations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
\begin{align}
  t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}
We note that $t_\mathrm{cool}$ does not explicitly depend on $\rho$.

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{\Sigma^2\Omega_k}{\eta\rho^2}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that $\beta$ depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv 
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
  \beta(z; \khat) =
  &\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\notag\\ 
&\times\left[\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right].         
\end{align} 
We refer to the case with $\gamma=1.4$, $\mu=2.33$ and the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN) and $\beta=\beta_\mathrm{MMSN}$, 
\begin{align}\label{beta_mmsn_simp}
  \beta_\mathrm{MMSN} = 3.9\times10^{-3} r_\mathrm{AU}^{9/14}\left( 1 +
    1.9\times10^7\hat{\rho}^2r_\mathrm{AU}^{-33/7}\khat^{-2}\right). 
\end{align}
% where
% \begin{align}
%   \
% %  \frac{3.7\times10^{-3}}{\mu\left(\gamma-1\right)\hat{\kappa}_d\hat{T}^5}, 
% \end{align}
% and
% \begin{align}
%   \khat_0 = 4.4\times10^{3}\hat{\kappa}_d\hat{T}^2\hat{\Sigma}.
% \end{align}
It is instructive to compare the above thermal timescale 
with the criterion Eq. \ref{iso_vsi_cond} evaluated for the MMSN,     
\begin{align}\label{bcrit_mmsn}
  % \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
  % -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
  \beta_\mathrm{crit,MMSN} = 0.024r_\mathrm{AU}^{2/7}. 
\end{align}

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ $\gamma=1.4$, $\hat{T}=\hat{\Sigma}=1$ for three
opacity scales: $\hat{\kappa}_d=0.1, \,1 $ (MMSN) and $10$ at two
heights: $z=0$ and $z=3H$. We also plot Eq. \ref{bcrit_mmsn} in each
panel.

Consider first the MMSN shown in the middle panels of 
Fig. \ref{mmsn_bcrit_bcool}. The critical thermal timescale,
Eq. \ref{bcrit_mmsn}, is useful in the optically thin regime (where the
curves overlap), since in this limit $\beta$ does not depend on
height, as assumed in our analysis. We thus infer that the VSI can
operate at tens of AU in the MMSN.

In the inner few AU of the MMSN, however, $\beta$ 
may be smaller or greater than $\beta_\mathrm{crit}$ depending on
$z$ as the disk transitions from an optically thick midplane to an
optically thin atmosphere. 
%Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
%the optically thick midplane, but with increasing height the disk 
%becomes optically thin and $\beta<\beta_\mathrm{crit}$. 
Our analytical criteria is not valid for variable $\beta$, and we must
turn to numerical solutions of the linear problem to determine whether
or not the VSI operates. We do this in the next section.

Lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}, shifts the curves inwards but also raises
them. In fact, lowering the opacity by an order of magnitude renders  
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This is 
expected to stabilize the disk. Raising the opacity, as shown in 
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the instability is
restricted to larger radii.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
for sufficiently large $\khat$. We
illustrate this in Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at 
each radius the value of $\khat$ such  
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI would only
occur on very small radial scales in the inner disk. % However, these
% may not be important as 
% our
% linear calculations suggest growth rates eventually decay for large
% $\khat$.

However, disturbances with large $\khat$ may be subject to viscous
decay in a real disk. The viscous timescale for a perturbation lengthscale 
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha_s\khat^2\Omega $, where
$\alpha_s$ is the standard alpha viscosity parameter
\citep{shakura73}. Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/\epsilon\Omega$ implies 
$\khat^2\lesssim \epsilon/\alpha_s$. Inserting $\epsilon \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha_s\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha_s\sim
10^{-5}$. \citep[These low viscosity levels are required for the
VSI, see][]{nelson13}.

@@

\subsection{VSI in the MMSN}
Here we solve the linear stability problem in the MMSN with the scale and
height dependent thermal relaxation timescale defined by
Eq. \ref{beta_mmsn_simp}.  We use a vertical 
domain $z_\mathrm{max}=5H$. In Fig. \ref{mmsn_overall} we plot growth
timescales $t_\mathrm{grow} \equiv 1/\nu$ associated with the
fundamental mode in units of the local orbital time
$P_\mathrm{orb}\equiv 2\pi/\Omega_k$.

In fact, for $\khat \lesssim 30$ we either only find the
fundamental mode, or that it is the most unstable, and the associated
growth rates are insensitive to $z_\mathrm{max}$. This is consistent with our
analytical discussion (\S\ref{iso_vsi_beta_crit}) and previous
numerical results (\S\ref{therm_relax_eff}), which indicate  
higher order modes are more easily stabilized by finite thermal
relaxation.

However, for large $\khat$ and/or $z_\mathrm{max}$, we also found
surface modes with shorter growth times than those shown in
Fig. \ref{mmsn_overall}. Such modes may not be robust because    
they are affected by boundary conditions (\S\ref{surf_comment}) and/or
viscosity in a real disk.

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits. It is necessary to
consider smaller scales as $r$ decreases as the disk generally becomes
optically thick.

Fig. \ref{mmsn_overall} is roughly consistent with the 
midplane thermal relaxation timescales for the MMSN ( 
Fig. \ref{mmsn_bcrit_bcool}, left middle panel). For example, for 
$\khat=10$ we find $\beta \gtrsim \beta_\mathrm{crit}$ for $r\lesssim
10$AU at $z=0$; and Fig. \ref{mmsn_overall} show stabilization inside
$\sim 8$AU where $t_\mathrm{grow}\to\infty$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the VSI in
    the MMSN. These are calculated using the scale and height
    dependent thermal relaxation given by Eq. \ref{beta_mmsn_simp}. 
    \label{mmsn_overall}}    
\end{figure}

@@

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with much
reduced growth rates (when scaled by the local 
rotation).

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column due to the vertically
global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests that the VSI can operate 
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer at $r=5$AU than at $r=50$AU: %  for which  $\beta <
% \beta_\mathrm{crit}  \forall z$: 
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU. Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_5AU}
  \caption{Vertical velocity perturbations of the VSI in the MMSN at 
    $r=50$AU (top) and $r=5$AU (bottom). The real (imaginary) part of 
    $\delta v_z$ is shown as
    the solid (dotted) curves. 
    \label{mmsn_eigenvz}}
\end{figure}
